---
title: "Seminar 5: Complexity Reduction, PCA, EFA, MDS, Clustering, Classification"
output: html_document
---

Purpose: learn PCA vs MDS, EFA, clusterization and classifier.  
Dataset: `datasets/ifood_df.csv`.  
Requires: dplyr, ggplot2, psych, randomForest, caret.  
Style: [adv-r.had.co.nz/Style.html](http://adv-r.had.co.nz/Style.html).

## Load data

```{r setup}
library(dplyr)
library(ggplot2)
library(randomForest)
library(psych)
library(caret)

df <- read.csv("datasets/ifood_df.csv", stringsAsFactors = FALSE)
set.seed(42)

# Exclude constant columns for PCA/MDS/EFA
constant_columns <- sapply(df, function(x) var(x) == 0)
df <- df[, !constant_columns]

# Numeric subset for EFA (used in Q2)
df_num <- df[, sapply(df, is.numeric)]
```

## Q1: PCA and MDS, comment on efficiency/relevance

PCA: captures variance in linear combinations. Efficient for high-dimensional numeric data and gives variance interpretation.  
MDS: preserves pairwise distances. Relevant when distance between observations is meaningful (e.g. perceptual mapping).

```{r q1-pca}
pca_fit <- prcomp(df, scale. = TRUE)
pca_scores <- as.data.frame(pca_fit$x)
summary(pca_fit)

# Variance explained by first few components
cumsum(pca_fit$sdev^2) / sum(pca_fit$sdev^2)
plot(pca_fit, type = "l")

# double check with a chart
ggplot(pca_scores, aes(x = PC1, y = PC2)) +
  geom_point() +
  labs(title = "Two Principles Components derived by the Elbow rule") +
  theme_minimal()
```

```{r q1-mds}
d <- dist(scale(df))
mds_fit <- cmdscale(d, k = 2)
mds_scores <- as.data.frame(mds_fit)

# double check with a chart
ggplot(mds_scores, aes(x = V1, y = V2)) +
  geom_point() +
  labs(title = "MDS for two dimensions")
```

## Q2: Exploratory Factor Analysis on numeric variables

```{r q2}
fa_cols <- c(
  "Income", "MntWines", "MntFruits", "NumDealsPurchases",
  "NumWebPurchases", "NumStorePurchases", "NumWebVisitsMonth"
)
fa_cols <- intersect(fa_cols, names(df_num))
df_fa <- df_num[, fa_cols]
df_fa <- df_fa[complete.cases(df_fa), ]
n_factors <- 2
efa_fit <- psych::fa(df_fa, nfactors = n_factors, rotate = "varimax", fm = "ml")
print(efa_fit)
```

Interpretation: loadings show which variables define each factor; e.g. factor 1 might be "total spending", factor 2 "channel usage".

## Q3: Cluster clients to predict/explain Income

Optimal (maybe): k-means on scaled numeric features. k-means because it is fast, scalable, and segment means are interpretable.

```{r q3}
df_clus <- scale(df)
k <- 3
km <- kmeans(df_clus, centers = k, nstart = 25)
df_clus_data <- df
df_clus_data$cluster <- factor(km$cluster)

# Mean Income by cluster (clustering "predicts" segment, Income varies by it)
# Not really consistent, but still
income_by_cluster <- aggregate(Income ~ cluster, data = df_clus_data, FUN = mean)
income_by_cluster

# double check by a chart
ggplot(df_clus_data, aes(x = Age, y = Kidhome, color = cluster)) +
  geom_point() +
  labs(title = "Clustering Results", x = "Age", y = "MntTotal") +
  theme_minimal()
```

## Q4: Classifier for Response (Accuracy and F1)

```{r q4-setup}
df_cls <- df
df_cls$Response_f <- factor(df_cls$Response)
preds <- c(
  "Income", "Recency", "NumDealsPurchases", "NumWebPurchases",
  "NumCatalogPurchases", "NumStorePurchases", "NumWebVisitsMonth",
  "MntWines", "MntTotal", "Age", "Kidhome"
)
preds <- intersect(preds, names(df_cls))
df_cls <- df_cls[complete.cases(df_cls[, c("Response_f", preds)]), ]

# train-test split
train_indices <- createDataPartition(df_cls$Response_f, p = 0.8, list = FALSE)
train_data <- df_cls[train_indices, ]
test_data <- df_cls[-train_indices, ]

# Function to check the final results
show_results <- function(test_dataframe, pred_test_data) {
  confusionMatrix <- confusionMatrix(pred_test_data, test_dataframe$Response_f)
  accuracy <- confusionMatrix$overall["Accuracy"]
  tp <- confusionMatrix$table[1, 1]
  fp <- confusionMatrix$table[2, 1]
  fn <- confusionMatrix$table[1, 2]
  precision_1 <- tp / (tp + fp)
  recall_1 <- tp / (tp + fn)
  f1_1 <- 2 * precision_1 * recall_1 / (precision_1 + recall_1)
  cat("Accuracy:", accuracy, "\n")
  cat("F1 (Response=1):", f1_1, "\n")
}
```

```{r q4-models}
# Can be done either with RandomForest
rf_cls <- randomForest(
  Response_f ~ .,
  data = train_data[, c("Response_f", preds)],
  ntree = 100
)
pred_test_rf <- predict(rf_cls, test_data)

# Or just GLM
glm_cls <- glm(Response_f ~ ., data = train_data[, c("Response_f", preds)],
               family = binomial)
pred_p <- predict(glm_cls, test_data, type = "response")
pred_test_glm <- factor(ifelse(pred_p > 0.5, 1, 0), levels = c(0, 1))

# More or less adequate accuracy and F1 (almost identical)
show_results(test_data, pred_test_rf)
show_results(test_data, pred_test_glm)
```
