---
title: "Seminar 3: Statistics Fundamentals"
output: html_document
---

Purpose: normality/homoscedasticity, t-test, factor merge, group tests, smart ANOVA/Kruskal-Wallis, Bayesian ANOVA.  
Dataset: `datasets/ifood_df.csv`.  
Requires: dplyr, tidyr, car (Levene), BayesFactor (optional).  
Style: [adv-r.had.co.nz/Style.html](http://adv-r.had.co.nz/Style.html).

## Load data and purchase columns

```{r setup}
library(dplyr)
library(tidyr)
library(car)

df <- read.csv("datasets/ifood_df.csv", stringsAsFactors = FALSE)

# Purchase columns for normality/homoscedasticity
purchase_cols <- colnames(select(df, ends_with("Purchases")))
```

## 1. Normality and homoscedasticity for purchase columns

Returns TRUE if any p-value < 0.05 (i.e. assumptions violated).

```{r norm-homosced}
norm_homosced_test <- function(dataframe) {
  cols <- intersect(purchase_cols, names(dataframe))
  if (length(cols) == 0) return(NA)

  # Normality: Shapiro-Wilk on a sample (max 5000) per column
  norm_pvals <- numeric(length(cols))
  set.seed(42)
  for (i in seq_along(cols)) {
    x <- dataframe[[cols[i]]][!is.na(dataframe[[cols[i]]])]
    x <- x[sample(length(x), min(5000, length(x)))]
    if (length(x) >= 3) {
      norm_pvals[i] <- shapiro.test(x)$p.value
    } else {
      norm_pvals[i] <- 1
    }
  }

  # Homoscedasticity: need a grouping variable - let's use Education as a variable
  long <- dataframe %>%
    select(all_of(cols)) %>%
    tidyr::pivot_longer(everything(), names_to = "variable", values_to = "value")
  long <- long[!is.na(long$value), ]
  if (nrow(long) < 10) return(TRUE)

  # Use variable as group for Levene (homogeneity across purchase types)
  tryCatch({
    lev <- car::leveneTest(value ~ Education, data = long)
    homo_pval <- lev[["Pr(>F)"]][1]
  }, error = function(e) {
    homo_pval <<- 1
  })
  if (!exists("homo_pval")) homo_pval <- 1

  return(any(c(norm_pvals, homo_pval) < 0.05))
}
norm_homosced_test(df)
```

## 2. One-sample t-test: group mean vs given mean

```{r compare-group}
compare_group_stat <- function(dataframe, column, given_mean) {
  x <- dataframe[[column]]
  x <- x[!is.na(x)]
  t.test(x, mu = given_mean)$p.value
}
compare_group_stat(df, "NumWebPurchases", 4)
```

## 3. Merge education_* columns into one factor

```{r gather-education}
gather_education <- function(dataframe) {
  edu_idx <- grep("^education_", names(dataframe))
  if (length(edu_idx) == 0) return(dataframe)

  edu_mat <- as.matrix(dataframe[, edu_idx, drop = FALSE])
  edu_names <- names(dataframe)[edu_idx]
  edu_level <- apply(edu_mat, 1, function(x) {
    w <- which(x == 1)
    if (length(w) == 0) NA else edu_names[w]
  })
  education <- gsub("^education_", "", edu_level)
  education <- factor(education)

  out <- dataframe[, -edu_idx, drop = FALSE]
  out$Education <- education
  return(out)
}
df_edu <- gather_education(df)
```

## 4. H0: No difference in Kidhome by Education

Use Kruskal-Wallis (robust to non-normality) or ANOVA.

```{r test-kidhome}
test_h0_kidhome <- function(dataframe) {
  d <- gather_education(dataframe)
  k <- kruskal.test(Kidhome ~ Education, data = d)
  return(k$p.value)
}
test_h0_kidhome(df_edu)
```

## 5. H0: No difference in NumWebPurchases by Kidhome

Kidhome is numeric - treat as factor for group comparison.

```{r test-numweb}
test_h0_numwebpurchases <- function(dataframe) {
  d <- dataframe
  d$Kidhome_f <- factor(d$Kidhome)
  k <- kruskal.test(NumWebPurchases ~ Kidhome_f, data = d)
  return(k$p.value)
}
test_h0_numwebpurchases(df)
```

## 6. ANOVA if assumptions met, else Kruskal-Wallis

Compare purchase columns across Education groups.

```{r smart-var-test}
smart_var_test <- function(dataframe, dependent, group) {
  x <- dataframe[[dependent]]

  # shapiro with sampling
  norm_p <- shapiro.test(x[sample(length(x), min(5000, length(x)))])$p.value
  homo_p <- bartlett.test(x ~ dataframe[[group]])$p.value
  if (norm_p > 0.05 & homo_p > 0.05) {
    test_result <- aov(as.formula(paste(dependent, "~", group)), data = dataframe)
    return(summary(test_result))
  } else {
    test_result <- kruskal.test(as.formula(paste(dependent, "~", group)),
                                data = dataframe)
    return(test_result)
  }
}
smart_var_test(df_edu, "NumStorePurchases", "Education")
```

## 7. Bayesian ANOVA on given numeric column

```{r bayes-anova}
bayes_anova <- function(dataframe, col, group) {
  dataframe$y <- dataframe[[col]]
  dataframe$group <- factor(dataframe[[group]])
  dataframe <- dataframe[!is.na(dataframe$y), ]
  BayesFactor::anovaBF(y ~ group, data = dataframe, progress = FALSE)
}
bayes_anova(df_edu, "NumStorePurchases", "Kidhome")
```
