---
title: "Seminar 4: Correlation and Regression (Midterm)"
output: html_document
---

Purpose: normality/homogeneity, group comparisons, linear/logistic/mixed models, optimal OLS, optional Random Forest.  
Dataset: `datasets/ifood_df.csv`.  
Requires: dplyr, car, caret, stringr, ggplot2, randomForest, lme4, MASS.  
Style: [adv-r.had.co.nz/Style.html](http://adv-r.had.co.nz/Style.html).

## Load data and helpers

```{r setup}
library(dplyr)
library(car)
library(caret)
library(stringr)
library(ggplot2)
library(randomForest)
library(lme4)
library(MASS)

df <- read.csv("datasets/ifood_df.csv", stringsAsFactors = FALSE)
set.seed(42)

# will need it for almost all tasks
df$Response_f <- factor(df$Response)

# helper: modified function from Seminar 3 to create Education/Marital factors
gather_factor <- function(dataframe, factor_name) {
  idx <- grep(paste(c("^", factor_name, "_"), collapse = ""), names(dataframe))
  if (length(idx) == 0) return(dataframe)

  mat <- as.matrix(dataframe[, idx, drop = FALSE])
  names <- names(dataframe)[idx]
  level <- apply(mat, 1, function(x) {
    w <- which(x == 1)
    if (length(w) == 0) NA else names[w]
  })
  final_factor <- gsub(paste(c("^", factor_name, "_"), collapse = ""), "", level)
  final_factor <- factor(final_factor)

  out <- dataframe[, -idx, drop = FALSE]
  out[[str_to_sentence(factor_name)]] <- final_factor
  return(out)
}
df <- gather_factor(df, "education")
df <- gather_factor(df, "marital")
```

## Q1: Normality and homogeneity for Income and MntWines

```{r q1}
income_ok <- df$Income[!is.na(df$Income)]
mntwines_ok <- df$MntWines[!is.na(df$MntWines)]

# Normality: Shapiro-Wilk with sampling
shapiro_income <- shapiro.test(sample(income_ok, min(5000, length(income_ok))))
shapiro_mntwines <- shapiro.test(sample(mntwines_ok, min(5000, length(mntwines_ok))))

# Homogeneity: Levene by Response for Income and MntWines
lev_income <- leveneTest(Income ~ Response_f, data = df)
lev_mntwines <- leveneTest(MntWines ~ Response_f, data = df)

# p < 0.05 suggests that none of the assumptions hold
shapiro_income$p.value
shapiro_mntwines$p.value
lev_income[["Pr(>F)"]][1]
lev_mntwines[["Pr(>F)"]][1]

# double check with charts
ggplot(df, aes(x = Income)) +
  geom_histogram(bins = 100, fill = "blue", alpha = 0.3) +
  ggtitle("Histogram of Income")
```

## Q2: Difference in Income by Response

Use Mann-Whitney since it is non-normal: reject the null hypothesis and say that there is a difference in Income between groups by Response.

```{r q2}
wilcox.test(Income ~ Response_f, data = df)
```

## Q3: Difference in NumDealsPurchases by Education

Since assumptions are violated - use Kruskal-Wallis test. Fail to reject the null hypothesis: there is not enough statistical evidence to conclude that there is a significant difference in the number of deals purchased between different education levels.

```{r q3}
kruskal.test(NumDealsPurchases ~ Education, data = df)

# double check with charts
ggplot(df, aes(x = Education, y = NumDealsPurchases, colour = Education)) +
  geom_boxplot() +
  labs(title = "Number of Deals Purchased by Education Level",
       x = "Education Level",
       y = "Number of Deals Purchased") +
  theme_minimal()
```

## Q4: Model to predict Income from factor variables

```{r q4}
# Transform all needed columns into factor type
factor_vars <- c(
  "Education", "Response", "Marital", "Complain", "Kidhome", "Teenhome"
)
df <- df %>%
  mutate_at(factor_vars, factor)

# Use these factors to construct a formula
formula <- as.formula(paste("Income ~", paste(factor_vars, collapse = " + ")))

# Summary of the model to predict Income
summary(lm(formula, data = df))
```

Interpretation: The most 'valuable' factors seem to be related to number of kids and education level. Overall adjusted R² = 0.36 is a pretty decent fit.

## Q5: Top-3 factors for Response (logistic)

```{r q5}
pred_cols <- names(df)[sapply(df, is.numeric)]
pred_cols <- intersect(pred_cols, names(df))
fmla_resp <- as.formula(paste("Response_f ~", paste(pred_cols, collapse = " + ")))
glm_resp <- glm(fmla_resp, data = df, family = binomial)
coef_tab <- summary(glm_resp)$coefficients
coef_tab <- coef_tab[rownames(coef_tab) != "(Intercept)", , drop = FALSE]
top3 <- head(coef_tab[order(abs(coef_tab[, "z value"]), decreasing = TRUE), ], 3)
top3

# also can do it with RF approach - interestingly results differ, though Recency is in both
rf_model <- randomForest(Response_f ~ . - `Response`, data = df, importance = TRUE)
importance_rf <- importance(rf_model)
importance_df <- data.frame(Variable = rownames(importance_rf), Importance = importance_rf[, 1])
importance_df <- importance_df %>% arrange(desc(Importance))

top_factors <- head(importance_df, 3)
top_factors
```

## Q6: Fixed and random effects of Education and Marital on MntTotal

```{r q6}
# Fixed-effects model with Education * Marital
lm_mnt <- lm(MntTotal ~ Education * Marital, data = df)
summary(lm_mnt)

# For true random effects (e.g. random intercept by customer segment)
fixed_md <- lmer(MntTotal ~ Education * Marital + (1 | Education) + (1 | Marital),
                data = df)
summary(fixed_md)
```

## Q7: Optimal model for NumStorePurchases (adjusted R²)

```{r q7}
train_indices <- createDataPartition(df$NumStorePurchases, p = 0.8, list = FALSE)
train_data <- df[train_indices, ]
test_data <- df[-train_indices, ]
lm_full <- lm(NumStorePurchases ~ ., data = train_data)
summary(lm_full)$adj.r.squared
summary(lm_full)

# Now removing non-significant predictors
lm_opt <- stepAIC(lm_full, direction = "both", trace = 0)
summary(lm_opt)
summary(lm_opt)$adj.r.squared  # small improvement

# Running optimized model on Test data
predictions <- predict(lm_opt, newdata = test_data)
model_performance <- postResample(predictions, test_data$NumStorePurchases)
r_squared <- cor(test_data$NumStorePurchases, predictions)^2
n <- nrow(test_data)
p <- length(coef(lm_opt)) - 1

# even better - 0.638, consistent with train results (also good)
adjusted_r_squared <- 1 - ((1 - r_squared) * (n - 1) / (n - p - 1))
adjusted_r_squared
```

## Q8 (*): Optimal model for Income (adjusted R²)

```{r q8}
lm_full_2 <- lm(Income ~ ., data = df)
lm_opt_2 <- stepAIC(lm_full_2, direction = "both", trace = 0)

# It seems like it is more reliable to predict Income from this dataset
summary(lm_opt_2)$adj.r.squared
```

## Q9 (*): Random Forest for AcceptedCmp1

```{r q9}
# make column a factor, exclude all AcceptedCmp*
df$AcceptedCmp1_f <- as.factor(df$AcceptedCmp1)
preds_rf <- names(df)[sapply(df, is.numeric)]
matches <- str_detect(preds_rf, "AcceptedCmp*")
preds_rf <- preds_rf[!matches]
preds_rf <- intersect(preds_rf, names(df))

# clean and do a train-test split
df_rf <- df[complete.cases(df[, c("AcceptedCmp1_f", preds_rf)]), ]
train_indices <- createDataPartition(df_rf$AcceptedCmp1_f, p = 0.8, list = FALSE)
train_data <- df_rf[train_indices, ]
test_data <- df_rf[-train_indices, ]

rf_fit <- randomForest(
  AcceptedCmp1_f ~ .,
  data = train_data[, c("AcceptedCmp1_f", preds_rf)],
  ntree = 100
)
rf_fit

# Confusion matrix, accuracy for Test data - pretty solid 93.6%; Bad specificity though
predictions <- as.factor(predict(rf_fit, newdata = test_data))
confusion_matrix <- confusionMatrix(predictions, test_data$AcceptedCmp1_f)
confusion_matrix
```
