---
title: "Seminar 1: Basics"
output: html_document
---

Purpose: descriptive statistics, central tendency, distributions, H0/errors.  
Dataset: `datasets/october_orders.csv`.  
Style: [adv-r.had.co.nz/Style.html](http://adv-r.had.co.nz/Style.html).

## Load data

```{r load-data}
library(dplyr)
library(ggplot2)

orders <- read.csv("datasets/october_orders.csv", stringsAsFactors = FALSE)
orders$lcl_dttm <- as.POSIXct(orders$lcl_dttm, format = "%Y-%m-%d %H:%M:%S")
```

## Q1: Central tendency for assignment_min

Compare mean, median, and skew to recommend the best measure.

```{r q1-central-tendency}
assignment_min <- orders$assignment_min[!is.na(orders$assignment_min)]
mean_assign <- mean(assignment_min)
median_assign <- median(assignment_min)
skew_approx <- (mean_assign - median_assign) / sd(assignment_min)

# According to a rule of thumb (-0.5 ... + 0.5), mean should be used in this case
if (abs(skew_approx) > 0.5) {
  best_central <- "median"
} else {
  best_central <- "mean"
}
best_central
```

But! for this scenario of right-skewed data, median will be more accurate (outliers pull the mean too much). We can see this clearly on chart below.

```{r q1-chart}
orders_filtered <- orders %>%
  filter(!is.na(assignment_min) & assignment_min <= 5)
ggplot(orders_filtered, aes(x = assignment_min)) +
  geom_histogram(fill = "lightblue") +
  geom_vline(xintercept = mean_assign, linetype = "dashed", color = "red") +
  annotate("text", x = mean_assign + 0.2, y = length(orders_filtered$assignment_min) / 3,
           label = "mean", color = "red") +
  geom_vline(xintercept = median_assign, linetype = "dashed", color = "green") +
  annotate("text", x = median_assign + 0.2, y = length(orders_filtered$assignment_min) / 3,
           label = "median", color = "green") +
  labs(title = "Assignment time distribution", y = "Frequency") +
  theme_minimal()
```

## Q2: Standard deviation of arrival_min

```{r q2}
sd_arrival_overall <- sd(orders$arrival_min, na.rm = TRUE)
sd_arrival_faster <- sd(orders$arrival_min[orders$tariff == "faster" &
  !is.na(orders$arrival_min)])

sd_arrival_overall
sd_arrival_faster
```

## Q3: Normality of order_fare

Shapiro-Wilk is sensitive to sample size; will use a subsample if n > 5000.

```{r q3}
order_fare_all <- orders$order_fare[orders$order_fare > 0 & !is.na(orders$order_fare)]
order_fare_cancel <- orders$order_fare[orders$status == "cancel" &
  !is.na(orders$order_fare)]
order_fare_cancel <- order_fare_cancel[order_fare_cancel >= 0]

set.seed(42)
n_sample <- min(5000, length(order_fare_all))
shapiro_all <- shapiro.test(sample(order_fare_all, n_sample))
# For cancel subset, run only if enough observations
shapiro_cancel <- if (length(order_fare_cancel) >= 20) {
  shapiro.test(sample(order_fare_cancel, n_sample))
} else {
  list(p.value = NA)
}

# p < 0.05 suggests deviation from normality
normality_all <- shapiro_all$p.value >= 0.05
# double check with a chart
hist(order_fare_all)

# order_fare for cancel is also not normal (right-skewed and many zeros, as expected)
normality_cancel <- is.na(shapiro_cancel$p.value) || shapiro_cancel$p.value >= 0.05
hist(order_fare_cancel)
```

## Q4: IQR for arrival_min

```{r q4}
iqr_arrival <- IQR(orders$arrival_min, na.rm = TRUE)
iqr_arrival

# double check with the chart
ggplot(orders, aes(y = arrival_min)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Arrival time distribution", y = "Frequency") +
  theme_minimal()
```

## Q5: H0 for difference in order_fare by status

H0: There is no significant difference in the mean (or distribution) of order_fare between different statuses (e.g. success vs cancel).

## Q6: Type I and Type II for H0 = no difference in assignment_min between last 2 weeks

- **Type I error (alpha):** Reject H0 when H0 is true. "We conclude there is a significant difference between the last 2 weeks when there isn't"
- **Type II error (beta):** Fail to reject H0 when H0 is false. "We conclude there is no significant difference when in fact there is"

## Q7: Mode for order_fare

Mode is the most frequently occurring value.

```{r q7}
order_fare_vec <- orders$order_fare[!is.na(orders$order_fare)]
freq <- table(order_fare_vec)
mode_order_fare <- as.numeric(names(freq)[which.max(freq)])
mode_order_fare

# Optional: function to compute mode for any numeric vector (using tabulate)
get_mode <- function(x) {
  uniqv <- unique(x)
  uniqv[which.max(tabulate(match(x, uniqv)))]
}
get_mode(orders$order_fare)
```
